Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-03-02 msc-publication.tex
remove 250m from gravity wave sampleLine plot, include test case setups and cut cell/snappyHexMesh description in paper

Much work has been done to reduce error associated with TF coordinates: firstly by smoothing the effects of terrain with height \citep{schaer2002,leuenberger-2010,klemp2011} and, secondly, by improving the accuracy in calculating the horizontal pressure gradient itself \citep{klemp2011,zaengl2012}.

@@

\section{Grids}
\TODO{introduce this section}
\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

\TODO{Explain how SnapCol grid is constructed?}

@@

Following \citet{schaer2002}, a tracer is transported above wave-shaped terrain by solving the advection equation for a prescribed horizontal wind.  This challenges the accuracy of the advection scheme in the presence of grid distortions.  The spatial domain, terrain profile, wind field, tracer, SLEVE scale heights are given by \citet{schaer2002}.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.  A tracer is placed above and to the left of the mountain with a magnitude of 1 at its centre and 0 at the perimeter.

While the continuous wind field specified by \citet{schaer2002} is non-divergent, this is not true of the discrete wind field on non-orthogonal grids.  A potential, $p$, is used to correct the wind field, $\bm{u}$, so that is non-divergent such that
\begin{align}
	\nabla \cdot \left( \bm{u} + \nabla p \right) = 0,
\intertext{which can be rearranged to give a Poisson equation}
	\nabla^2 p = - \nabla \cdot \bm{u}.
\end{align}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) the fourth-order centred difference scheme with sigma coordinates from \citet{schaer2002}, and the upwind-biased cubic scheme on (b) BTF grid, (c) cut cell grid.  Contour intervals are every 0.1.}
	\label{fig:advection-tracer}
\end{figure}

@@

\begin{table}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the tracer advection test.  Results of the cubic upwind-biased scheme are compared with the fourth-order centred scheme from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ l l l l l l }
\hline\hline
& \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{2}{c}{Sch\"ar 4th order} \\
& \(\ell^2\) error & min & max & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 1 \\
BTF 	  & \num{0.00758} & \num{-0.034} & \num{0.928} & \num{-0.058} & \num{1.001} \\
SLEVE 	  & \num{0.00110} & \num{-0.011} & \num{0.981} & \num{-0.002} & \num{0.984} \\
Cut cell  & \num{0.00058} & \num{-0.007} & \num{0.982} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
No orography & \num{0.00058} & \num{-0.007} & \num{0.982} & \num{-0.002} & \num{0.984} \\
\hline
\end{tabular}
\end{table}

\subsection{Resting atmosphere}
An idealised terrain profile is defined along with a stratified atmosphere in hydrostatic balance.  The analytic solution is time-invariant, but numerical errors in calculating the horizontal pressure gradient can give rise to spurious velocities which become more severe over steeper terrain \citet{klemp2011}.  The test setup follows the specification by \citet{klemp2011} but uses the smaller domain and rigid boundaries from \cite{weller-shahrokhi2014}.

@@

\subsection{Gravity waves}
\begin{itemize}
	\item \TODO{Motivation: it models a real dynamic process?}
	\item Specify domain, thermodynamics, prescribed inlet wind
	\item Specify sponge layers, BCs
	\item Compare BTF, SLEVE and SnapCol results (SLEVE only briefly as visually identical to BTF)
\end{itemize}

Analysis:
\begin{itemize}
	\item \(w\) contours visually similar on all grids, agree with \citet{melvin2010} (Figure~\ref{fig:gw-w})
	\item \TODO{divergence might help explain computational mode -- it's greater on the SnapCol grid (MSc dissertation Figure 4.15e)}
	\item \(\theta\) anomalies similar on all grids EXCEPT...
	\item ... on SnapCol grid in lee of mountain near the ground. (Figure~\ref{fig:gw-theta})
	\item \TODO{worth mentioning implications of this?  1. reduced stability, although not enough to create vertical motion in this instance 2. would disrupt clouds 2. model has no viscosity so thermal mixing should not occur in theory}
	\item Lorenz computation mode has been excited because Exner sample line at $x = \SI{50}{\kilo\meter}$ is in hydrostatic balance (Figure~\ref{fig:gw-exner-theta})
	\item \TODO{We can speculate on what excites the computational mode, but perhaps better not to?}
	\item No evidence of small cell problem -- \TODO{don't know that we can say much here without another vertical momentum test; our hypothesis about the quasi-horizontal flow isn't backed up with a test case}
	\item Conclusion: results similar on all grids, agree with literature, except for Lorenz computational mode on SnapCol grid
\end{itemize}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves-sampleLine/fig-gravityWaves-sampleLine.pdf}
%
	\caption{Vertical profiles of the Exner function of pressure, \(\Pi\), and potential temperature, \(\theta\), in the gravity waves test.  Exner profile is visually identical on all grids for both mountain heights; for clarity, the Exner profile is only plotted for the BTF grid.  The computational mode is manifested as a zig-zag in potential temperature on the cut cell grid which.  Doubling the mountain height from \SI{250}{\meter} to \SI{500}{\meter} increases the severity of the computational mode but has negligible effect on the BTF grid.  Results on the SLEVE grid (not shown) are qualitatively the same as those on the BTF grid for both mountain heights.}
	\label{fig:gw-exner-theta}
\end{figure}

>>>>

Much work has been done to reduce error associated with TF coordinates: firstly by smoothing the effects of terrain with height \citep{schaer2002,leuenberger2010,klemp2011} and, secondly, by improving the accuracy in calculating the horizontal pressure gradient itself \citep{klemp2011,zaengl2012}.

@@

\section{Grids}
\label{sec:grid}

\begin{figure*}
	\centering
	TODO
	%
	\caption{Examples of (a) BTF, (b) SLEVE, and (c) a cut cell grid, all constructed with OpenFOAM.  The two dimensional grids are \SI{20}{\kilo\meter} wide and \SI{2}{\kilo\meter} high.  The cut cell grid was created by intersecting the terrain surface with a regular grid as described in section~\ref{sec:grid}.}
	\label{fig:grid}
\end{figure*}

\TODO{introduce this section}
\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

Most implementations of terrain following layers use a coordinate system that makes the domain rectangular, but introduces metric terms into the equations of motion.  Instead, the model employed in this study uses Cartesian coordinates and unstructured grids.  By doing so, results from the same model can be compared between terrain following and cut cell grids without modifying the equation set or discretisation.

The OpenFOAM utility \texttt{snappyHexMesh} was used to create a grid that approximates the cut cell method.  First, a custom utility was used to move points beneath the surface up to the surface creating small cells near mountain peaks.  Second, the surface faces were taken from the BTF grid and \texttt{snappyHexMesh} was used to intersect the surface with the grid.  The tool removes cells whose centres are below the surface.  An example of the resulting grid is shown in figure~\ref{fig:grid}c.

There are two details of grid construction which mean that resulting cut cell grids can differ slightly from a typical cut cell grid created using a shaving method, as described by \citet{adcroft1997}.  First, when \texttt{snappyHexMesh} moves points along the surface according to its heuristics, some points are moved horizontally.  Second, the utility does not create new points necessary for pentagonal cells.

@@

Following \citet{schaer2002}, a tracer is transported above wave-shaped terrain by solving the advection equation for a prescribed horizontal wind.  This challenges the accuracy of the advection scheme in the presence of grid distortions.

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high which is discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
	h(x) &= \cos^2 \left( \frac{\pi x}{\lambda} \right) h^\star
%
	\intertext{where}
%
	h^\star(x) &= \left\{ \begin{array}{l l}
		h_0 \cos^2 \left( \frac{\pi x}{2a} \right) & \quad \text{if $| x | < a$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, and $\lambda = \SI{8}{\kilo\meter}$ is the wavelength.  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

The wind is entirely horizontal and is prescribed as
\begin{align}
	u(z) = u_0 \left\{ \begin{array}{l l}
		1 & \quad \text{if $z \geq z_2$} \\
		\sin^2 \left( \frac{\pi}{2} \frac{z - z_1}{z_2 - z_1} \right) & \quad \text{if $z_1 < z < z_2$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.	
\end{align}
where $u_0 = \SI{10}{\meter\per\second}$, $z_1 = \SI{4}{\kilo\meter}$ and $z_2 = \SI{5}{\kilo\meter}$.
This results in a constant wind aloft, and zero flow at \SI{4}{\kilo\meter} and below.

While the continuous wind field is non-divergent, this is not true of the discrete wind field on non-orthogonal grids.  A potential, $p$, is used to correct the wind field, $\bm{u}$, so that is non-divergent such that
\begin{align}
	\nabla \cdot \left( \bm{u} + \nabla p \right) = 0,
\intertext{which can be rearranged to give a Poisson equation}
	\nabla^2 p = - \nabla \cdot \bm{u}.
\end{align}

A tracer $\varphi$ is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\varphi(x, z) &= \varphi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius $r$ given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\varphi_0 = 1$ is the maximum magnitude of the anomaly.  At $t = \SI{0}{\second}$, the anomaly is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the anomaly is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by adjusting the anomaly centre such that $x_0 = ut$.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) the fourth-order centred difference scheme with sigma coordinates from \citet{schaer2002}, and the upwind-biased cubic scheme on (b) BTF grid, (c) cut cell grid.  Contour intervals are every 0.1.  The terrain profile is also shown immediately above the ground.}
	\label{fig:advection-tracer}
\end{figure}

@@

\begin{table}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the tracer advection test.  Results of the cubic upwind-biased scheme are compared with the fourth-order centred scheme from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{ l l l l l l }
\hline\hline
& \multicolumn{3}{c}{Cubic upwind-biased} & \multicolumn{2}{c}{Sch\"ar 4th order} \\
& \(\ell^2\) error & min & max & min & max \\
\hline
Analytic  & 0 & 0 & 1 & 0 & 1 \\
BTF 	  & \num{-1} & \num{-1} & \num{-1} & \num{-0.058} & \num{1.001} \\
SLEVE 	  & \num{-1} & \num{-1} & \num{-1} & \num{-0.002} & \num{0.984} \\
Cut cell  & \num{-1} & \num{-1} & \num{-1} & \multicolumn{1}{c}{---} & \multicolumn{1}{c}{---} \\
No orography & \num{-1} & \num{-1} & \num{-1} & \num{-0.002} & \num{0.984} \\
\hline
\end{tabular}
\end{table}

\subsection{Resting atmosphere}
An idealised terrain profile is defined along with a stratified atmosphere at rest in hydrostatic balance.  The analytic solution is time-invariant, but numerical errors in calculating the horizontal pressure gradient can give rise to spurious velocities which become more severe over steeper terrain \citet{klemp2011}.

The test setup follows the specification by \cite{klemp2011}, but has a narrower domain that is \SI{20}{\kilo\meter} wide and \SI{20}{\kilo\meter} high in order to reduce simulation time, as used by \citet{weller-shahrokhi2014}.  The grid resolution is \(\Delta x = \Delta z^\star = \SI{500}{\meter}\) as originally specified by \citet{klemp2011}.  All boundary conditions are no normal flow.

The wave-shaped mountain profile has a surface height, $h$, given by
\begin{align}
	h(x) = h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right) \cos^2 \left( \frac{\pi x}{\lambda} \right) \label{eqn:resting:mountain}
\end{align}
where $a = \SI{5}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{1}{\kilo\meter}$ is the maximum mountain height and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  For the optimised SLEVE grid, the large-scale component $h_1$ is specified as
\begin{align}
h_1(x) = \frac{1}{2} h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right)
\end{align}
and, following \cite{leuenberger2010}, $s_1 = \SI{4}{\kilo\meter}$ is the large scale height, $s_2 = \SI{1}{\kilo\meter}$ is the small scale height, and the optimal exponent value of $n = 1.35$ is used.

The initial thermodynamic conditions are in discrete hydrostatic balance, having a surface temperature of $\theta_0 = \SI{288}{\kelvin}$ and constant stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$ between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

@@

\subsection{Gravity waves}
\TODO{introduce and motivate the test... it models a real dynamic process}

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.

The initial thermodynamic conditions have a surface temperature of $\theta_0 = \SI{288}{\kelvin}$ and constant stability with $N = \SI{0.01}{\per\second}$ everywhere.  A constant horizontal wind $u = \SI{10}{\meter\per\second}$ is prescribed at the inlet boundary.

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The term $\mu \rho \bf{u}$ is subtracted from the momentum equation \TODO{include the equation set in this paper, or cite weller-shahrokhi2014 here?} where the damping function $\mu$ is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only.

Note that, while the domain itself is \SI{30}{\kilo\meter} in height, for the purposes of generating of BTF and SLEVE grids, the domain height is set to \SI{20}{\kilo\meter} because the sponge layer occupies the uppermost \SI{10}{\kilo\meter}.

No normal flow is imposed at the top and bottom boundaries and the outlet is zero gradient.  For Exner, hydrostatic balance is prescribed on all boundaries.  The simulation is integrated forward by 5 hours with a timestep $\Delta t = \SI{8}{\second}$.

Analysis:
\TODO{We compare BTF, SLEVE and SnapCol results (SLEVE only briefly as visually identical to BTF)}
\begin{itemize}
	\item \(w\) contours visually similar on all grids, agree with \citet{melvin2010} (Figure~\ref{fig:gw-w})
	\item \TODO{divergence might help explain computational mode -- it's greater on the SnapCol grid (MSc dissertation Figure 4.15e)}
	\item \(\theta\) anomalies similar on all grids EXCEPT...
	\item ... on SnapCol grid in lee of mountain near the ground. (Figure~\ref{fig:gw-theta})
	\item \TODO{worth mentioning implications of this?  1. reduced stability, although not enough to create vertical motion in this instance 2. would disrupt clouds 2. model has no viscosity so thermal mixing should not occur in theory}
	\item Lorenz computation mode has been excited because Exner sample line at $x = \SI{50}{\kilo\meter}$ is in hydrostatic balance (Figure~\ref{fig:gw-exner-theta})
	\item \TODO{We can speculate on what excites the computational mode, but perhaps better not to?}
	\item No evidence of small cell problem -- \TODO{don't know that we can say much here without another vertical momentum test; our hypothesis about the quasi-horizontal flow isn't backed up with a test case}
	\item Conclusion: results similar on all grids, agree with literature, except for Lorenz computational mode on SnapCol grid
\end{itemize}

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves-sampleLine/fig-gravityWaves-sampleLine.pdf}
%
	\caption{Vertical profiles of the Exner function of pressure, \(\Pi\), and potential temperature, \(\theta\), in the gravity waves test with a mountain height of \(h_0 = \SI{500}{\meter}\).  Exner profile is visually identical on all grids for both mountain heights; for clarity, the Exner profile is only plotted for the BTF grid.  The computational mode is manifested as a zig-zag in potential temperature on the cut cell grid which.   Results on the SLEVE grid (not shown) are qualitatively the same as those on the BTF grid.}
	\label{fig:gw-exner-theta}
\end{figure}
